Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-26 manuscript.tex
introduction

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Why do neurons communicate through spikes? By definition, a spike is an all-or-none event which occurs at continuous times. In other words, spikes are on one side binary, they occur or not without further details,  and second can occur at any asynchronous time, without a centralized clock. This stands in stark contrast to the discretized timing classically used in digital processing at the base of modern-day neural networks. In the living world, neural systems almost systematically use this so-called event-based representation. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging topic in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal mission in the age of the digital economy and of global warming. In this review paper, we provide evidence from biology and engineering that the precise timing of spikes plays a key role for our understanding of efficient neural networks.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
 Neural activity is directly influenced by our immediate environment and by internal states in order to generate actions. These are essential for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of neural activity. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality. In this respect, it is known that this neural activity is related to a specific variable, and since Galvani's experiments, we know that electrical activity can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). Among various phenomenon, a well studied way of communication between neurons are action potentials, or spikes. In the scope of this article we focus mainly on spikes, which are brief (about one millisecond) electrochemical impulses that propagate along the axons of neurons. These have the particularity of being essentially binary in their amplitude (i.e. they are prototypical, all or nothing).  The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. Neurophysiologists typically use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.  However, computational neuroscience models have suggested that precise timing may also play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{paugam-moisy_computing_2012}. We will investigate this very hypothesis in this review.

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Why do neurons communicate through spikes? By definition, a spike is an all-or-none event which occurs at continuous times. In other words, spikes are on one side binary, they occur or not without further details,  and second can occur at any asynchronous time, without a centralized clock. This stands in stark contrast to the discretized timing classically used in digital processing at the base of modern-day neural networks. In the living world, neural systems almost systematically use this so-called event-based representation. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging topic in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal mission in the age of the digital economy and of global warming. In this review paper, we provide evidence from biology and engineering that the precise timing of spikes plays a crucial role in our understanding of the efficiency of neural networks.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
 Neural activity is directly influenced by our immediate environment and by internal states, and is shaped in order to generate actions. The efficiency of these actions is key for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of neural activity, as one can observe it in neurobilogical experiments. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality. In this respect, it is known that neural activity may be related to a specific measurable variable, and since Galvani's experiments, we know that an electrical activity applied on musuclar nerves can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). Among various phenomenon, a well studied way of communication between neurons are specific electrochemical events called action potentials, or spikes. Typically, these are received by a neuron from other afferent neurons on the arborized "input" dendrite, integrated in its soma, and eventually further propagated along its "ouput" along the axon to reach efferent neurons. In the scope of this article, we focus mainly on these spikes, which are brief (about one millisecond) impulses that propagate along the axons of neurons. First, these have the particular property of being essentially binary in their amplitude (i.e. they are prototypical, all or nothing). The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. Neurophysiologists typically use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.  However, computational neuroscience models have suggested that precise timing may also play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{paugam-moisy_computing_2012}. We will investigate this very hypothesis in this review.
